### 2022-03-16 paper/localpsf.tex
restructuring paper

The output of our method is a hierarchical matrix (H-matrix) approximation, $\mathbf{A}_H$, of a discretization of $\Aop$. While operator actions of $\Aop$ on vectors are typically computationally expensive, fast H-matrix methods permit us to perform matrix-vector products with $\mathbf{A}_H$ cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving $\mathbf{A}_H$ that cannot easily be done using $\Aop$, because $\Aop$ is only available matrix-free. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

\section{Key concepts}
\label{sec:prerequisites}

In this section we present the key concepts that our method is based on. The method itself will be presented in Section \ref{sec:method}. First, we define moments of the impulse responses, $\phi_x$, and show how these moments can be computed efficiently (Section \ref{sec:intromoments}). Our method will use these moments extensively. Second, we present ellipsoid shaped a-priori estimates for the supports of the impulse responses (Section \ref{eq:mean_and_covariance_estimation}). Our method will use these ellipsoid estimates to choose sets of impulse responses that can be computed in batches. Third, we describe an impulse response approximation that we call ``normalized local mean displacement invariance'' (Section \ref{sec:local_mean_displacement_invariance}). Our method will interpolate impulse responses using this approximation. Fourth, we discuss how discretization of continuous functions by finite elements leads to matrix discretizations $\mathbf{A}$ and $\mathbf{\Phi}$ of $\Aop$ and $\Phi$, respectively (Section \ref{sec:finite_element_kernel}). The primary function of our method is to compute H-matrix approximations of these matrices. 
%While we use finite elements, other discretization schemes such as finite differences or finite volumes could be used instead with minor modifications.

@@

%The ellipsoid $E_x$ is not immediately available. 
%(recall $\gdim$ is the spatial dimension of $\Omega$).
The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because it requires one operator action of $\Aop$ per point $x$, and our method will need to know $V(x)$, $\mu(x)$, and $\Sigma(x)$ at all points $x$. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions. Specifically, let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^\gdim\right)$, and let $C$, $\{L^i\}_{i=1}^\gdim$, and ${\{Q^{ij}\}_{i=1}^\gdim}_{j=1}^\gdim$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) = x^i x^j.
\end{equation*}
The following results (derived in Appendix \ref{app:proofs}) follow straightforwardly from the definitions:
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i=1,\dots, \gdim$, $j=1,\dots,\gdim$. Here $f/g$ denotes pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denotes pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$.

@@

%We define the finite element approximation of the kernel, $\Phi^h \approx \Phi$, as follows:
The interpolation of $\Phi$ onto $V_h \otimes V_h$, which we denote by $\Phi^h$, is given as follows:
\begin{equation}
\label{eq:defn_of_Akerpcmesh}
\Phi(y,x) \approx \Phi^h(y,x) := \sum_{i=1}^\fedim \sum_{j=1}^\fedim \Phi(p_i,p_j) \febasis_i(y) \febasis_j(x).
\end{equation}
%The kernel $\Phi^h$ is the interpolation of $\Phi$ onto $V_h \otimes V_h$. Error in the overall kernel approximation, $\widetilde{\Phi}^h \approx \Aker$, arises both because of error in the product-convolution approximation, and because of finite element discretization error incurred from interpolation onto $V_h \otimes V_h$. 
Replacing $\Aker$ with $\Phi^h$ within the definition of $\Aop$ in \eqref{eq:kernel_representation}, then performing straightforward algebraic manipulations (see Appendix \ref{app:proofs} for details), yields the following approximation:
\begin{equation}
\label{eq:kernel_representation_pcmesh}
(\Aop u_h)(v_h) \approx \int_\Omega \int_\Omega v_h(y) \Phi^h(y,x) u_h(x) dx dy
= \bm{v}^T \bm{M} \mathbf{\Phi} \bm{M} \bm{u}.
\end{equation}
%While $\Aop$ is well-defined as a mapping $L^2(\Omega) \rightarrow L^2(\Omega)'$, we are interested in the restriction of $\Aop$ to functions in the finite element space. 
where $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ is the dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi(p_i, p_j),
\end{equation}
and $\bm{M} \in \mathbb{R}^{\fedim \times \fedim}$ is the finite element mass matrix which has entries
\begin{equation*}
\bm{M}_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx.
\end{equation*}
%Matrix entries $\Phi(p_i, p_j)$ are expensive to compute because $\Aop$ is only available matrix-free.
%Using \eqref{eq:defn_of_Akerpcmesh}, \eqref{eq:kernel_representation_pcmesh}, \eqref{eq:fem_coeff_basis}, and \eqref{eq:Akerpcmat_entries}, the definition of the mass matrix, and performing algebraic manipulations, it is straightforward but tedious to show that
%\begin{equation*}
%	\left(\AopPcMesh u_h \right)(v_h) = \bm{v}^T \bm{M} \AkerPcMat \bm{M} \bm{u}.
%\end{equation*}
Our matrix representation of $\Aop$ is therefore given by
\begin{equation}
\label{eq:Amatpc_defn}
\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Note that $\mathbf{A}$ is not the Galerkin projection of $\Aop$ onto $V_h$ because of the approximation $\Phi^h \approx \Phi$.
%because of the approximation $\Phi^h \approx \Phi$. 
Hence, $\mathbf{A}$ may differ slightly from the implicitly defined matrix that performs the mapping $\mathbf{u} \rightarrow \mathbf{v}$ corresponding to the operator action $v=\Aop u$.

@@

The results from Section \ref{sec:intromoments} allow us to compute $V$, $\mu$, and $\Sigma$ by applying $\Aop^T$ to a small number of polynomial functions, and the results of Section \ref{eq:mean_and_covariance_estimation} permit us to form an ellipsoid shaped estimate for the support of $\phi_x$, \emph{without} computing $\phi_x$. This allows us to compute large numbers of impulse responses, $\impulseresponse_{x_i}$, in ``batches,'' $\eta_b$ (see Figure \ref{fig:batches_intro}). We compute the impulse response batch $\eta_b$ by applying $\Aop$ to a weighted sum of point sources (Dirac comb) associated with a batch, $S_b$, of points $x_i$ scattered throughout $\Omega$ (Section \ref{sec:get_impulse_response}). The batch of points $S_b$ is chosen via a greedy ellipsoid packing algorithm so that, for $x_i,x_j \in S_b$, the support of $\impulseresponse_{x_i}$ and the support of $\impulseresponse_{x_j}$ do not overlap (or do not overlap much) if $i \neq j$ (Section \ref{sec:sample_point_selection}). Because these supports do not overlap, we can post-process $\eta_b$ to recover the functions $\impulseresponse_{x_i}$ associated with all points $x_i \in S_b$---with one application of $\Aop$, we recover many $\impulseresponse_{x_i}$ (Section \ref{sec:get_impulse_response}). The process is repeated to get more batches, until a desired number of impulse responses is reached.

@@

>>>>

The output of our method is a hierarchical matrix (H-matrix) approximation, $\mathbf{A}_H$, of a discretization of $\Aop$. H-matrices are a matrix format in which the matrix is re-ordered then recursively subdivided into blocks, so that many off-diagonal blocks of the matrix are low rank. For more details about H-matrices, see Appendix \ref{app:h_matrix}. While operator actions of $\Aop$ on vectors are typically computationally expensive, fast H-matrix methods permit us to perform matrix-vector products with $\mathbf{A}_H$ cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving $\mathbf{A}_H$ that cannot easily be done using $\Aop$, because $\Aop$ is only available matrix-free. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

@@

\section{Main concepts}
\label{sec:prerequisites}

In this section we present the three main concepts that our method is based on. The method itself will be presented in Section \ref{sec:method}. First, we define moments of the impulse responses, $\phi_x$, and show how these moments can be computed efficiently (Section \ref{sec:intromoments}). Our method will use these moments extensively. Second, we present ellipsoid shaped a-priori estimates for the supports of the impulse responses (Section \ref{eq:mean_and_covariance_estimation}). Our method will use these ellipsoid estimates to choose sets of impulse responses that can be computed in batches. Third, we describe an impulse response approximation that we call ``normalized local mean displacement invariance'' (Section \ref{sec:local_mean_displacement_invariance}). Our method will interpolate impulse responses using this approximation.

@@

%The ellipsoid $E_x$ is not immediately available. 
%(recall $\gdim$ is the spatial dimension of $\Omega$).
The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because it requires one operator action of $\Aop$ per point $x$, and our method will need to know $V(x)$, $\mu(x)$, and $\Sigma(x)$ at all points $x$. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions. Specifically, let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^\gdim\right)$, and let $C$, $\{L^i\}_{i=1}^\gdim$, and ${\{Q^{ij}\}_{i=1}^\gdim}_{j=1}^\gdim$ be the following constant, linear, and quadratic functions:
\begin{equation*}
C(x) := 1, \qquad
L^i(x) := x^i, \qquad
Q^{ij}(x) = x^i x^j.
\end{equation*}
The following results (derived in Appendix \ref{app:proofs}) follow straightforwardly from the definitions:
\begin{subequations}
	\label{eq:vol_mean_var}
	\begin{align}
	\spatialvol =& \left(\Aop^T C\right)^* \\
	\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
	\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
	\end{align}
\end{subequations}
for $i=1,\dots, \gdim$, $j=1,\dots,\gdim$. Here $f/g$ denotes pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denotes pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The resulting method form computing $V$, $\mu$, and $\Sigma$ is shown in Algorithm \ref{alg:varhpi_mean_cov}.

@@

@@

After discretization with finite elements\footnote{While we use finite elements, other discretization schemes such as finite differences or finite volumes could be used instead, with minor modifications to our method.} (Section \ref{sec:finite_element_kernel}), $\Aop$ and $\Phi$ become matrices $\mathbf{A}$ and $\mathbf{\Phi}$, respectively. Because $\Aop$ is only available matrix-free and $\Phi$ is only defined implicitly in terms of $\Aop$, the matrices $\mathbf{A}$ and $\mathbf{\Phi}$ are not easily accessible. Directly computing and storing these matrices is computationally intractible. In this section, we present our method for constructing accessible H-matrix approximations, $\mathbf{A}_H$ and $\mathbf{\Phi}_H$ of $\mathbf{A}$ and $\mathbf{\Phi}$, respectively, using a small number of operator actions of $\Aop$ and $\Aop^T$.

%Fourth, we discuss how discretization of continuous functions by finite elements leads to matrix discretizations $\mathbf{A}$ and $\mathbf{\Phi}$ of $\Aop$ and $\Phi$, respectively (Section \ref{sec:finite_element_kernel}). The primary function of our method is to compute H-matrix approximations of these matrices. 
%While we use finite elements, other discretization schemes such as finite differences or finite volumes could be used instead with minor modifications.

\subsection{Overview}
\label{sec:method_overview}

Algorithm \ref{alg:varhpi_mean_cov} from Section \ref{sec:intromoments} allows us to compute $V$, $\mu$, and $\Sigma$ by applying $\Aop^T$ to a small number of polynomial functions, and \eqref{eq:support_ellipsoid} from Section \ref{eq:mean_and_covariance_estimation} uses $V$, $\mu$, and $\Sigma$ to construct an ellipsoid shaped estimate for the support of $\phi_x$, \emph{without} computing $\phi_x$. This allows us to compute large numbers of impulse responses, $\impulseresponse_{x_i}$, in ``batches,'' $\eta_b$ (see Figure \ref{fig:batches_intro}). We compute the impulse response batch $\eta_b$ by applying $\Aop$ to a weighted sum of point sources (Dirac comb) associated with a batch, $S_b$, of points $x_i$ scattered throughout $\Omega$ (Section \ref{sec:get_impulse_response}). The batch of points $S_b$ is chosen via a greedy ellipsoid packing algorithm so that, for $x_i,x_j \in S_b$, the support of $\impulseresponse_{x_i}$ and the support of $\impulseresponse_{x_j}$ do not overlap (or do not overlap much) if $i \neq j$ (Section \ref{sec:sample_point_selection}). Because these supports do not overlap, we can post-process $\eta_b$ to recover the functions $\impulseresponse_{x_i}$ associated with all points $x_i \in S_b$---with one application of $\Aop$, we recover many $\impulseresponse_{x_i}$ (Section \ref{sec:get_impulse_response}). The process is repeated to get more batches, until a desired number of impulse responses is reached.

@@

%We define the finite element approximation of the kernel, $\Phi^h \approx \Phi$, as follows:
The interpolation of $\Phi$ onto $V_h \otimes V_h$, which we denote by $\Phi^h$, is given as follows:
\begin{equation}
\label{eq:defn_of_Akerpcmesh}
\Phi(y,x) \approx \Phi^h(y,x) := \sum_{i=1}^\fedim \sum_{j=1}^\fedim \Phi(p_i,p_j) \febasis_i(y) \febasis_j(x).
\end{equation}
%The kernel $\Phi^h$ is the interpolation of $\Phi$ onto $V_h \otimes V_h$. Error in the overall kernel approximation, $\widetilde{\Phi}^h \approx \Aker$, arises both because of error in the product-convolution approximation, and because of finite element discretization error incurred from interpolation onto $V_h \otimes V_h$. 
Replacing $\Aker$ with $\Phi^h$ within the definition of $\Aop$ in \eqref{eq:kernel_representation}, then performing straightforward algebraic manipulations (see Appendix \ref{app:proofs} for details), yields the following approximation:
\begin{equation}
\label{eq:kernel_representation_pcmesh}
(\Aop u_h)(v_h) \approx \int_\Omega \int_\Omega v_h(y) \Phi^h(y,x) u_h(x) dx dy
= \bm{v}^T \bm{M} \mathbf{\Phi} \bm{M} \bm{u}.
\end{equation}
%While $\Aop$ is well-defined as a mapping $L^2(\Omega) \rightarrow L^2(\Omega)'$, we are interested in the restriction of $\Aop$ to functions in the finite element space. 
where $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ is the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi(p_i, p_j),
\end{equation}
and $\bm{M} \in \mathbb{R}^{\fedim \times \fedim}$ is the sparse finite element mass matrix which has entries
\begin{equation*}
\bm{M}_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx.
\end{equation*}
%Matrix entries $\Phi(p_i, p_j)$ are expensive to compute because $\Aop$ is only available matrix-free.
%Using \eqref{eq:defn_of_Akerpcmesh}, \eqref{eq:kernel_representation_pcmesh}, \eqref{eq:fem_coeff_basis}, and \eqref{eq:Akerpcmat_entries}, the definition of the mass matrix, and performing algebraic manipulations, it is straightforward but tedious to show that
%\begin{equation*}
%	\left(\AopPcMesh u_h \right)(v_h) = \bm{v}^T \bm{M} \AkerPcMat \bm{M} \bm{u}.
%\end{equation*}
Our matrix representation, $\mathbf{A}$, of $\Aop$ is therefore given by
\begin{equation}
\label{eq:Amatpc_defn}
\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Note that $\mathbf{A}$ is not exactly the Galerkin projection of $\Aop$ onto $V_h$ because of the approximation $\Phi^h \approx \Phi$.
%because of the approximation $\Phi^h \approx \Phi$. 
Hence, $\mathbf{A}$ may differ slightly from the implicitly defined matrix that performs the mapping $\mathbf{u} \rightarrow \mathbf{v}$ corresponding to the discrete version of the operator action $v=\Aop u$. 
%It is computationally intractible to build and store the matrices $\mathbf{A}$ and $\mathbf{\Phi}$, and individual matrix entries $\mathbf{A}_{ij}$ and $\mathbf{\Phi}_{ij}$ are not easily accessible. In the remainder of this section, we show how to build accessible approximations of these matrices.
